### 2014-10-15 Paper_v1.tex
Working on discussion

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, and the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
However, ODEs often come with difficult (and sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers have shown that model behavior and predictive ability is typically 
sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions from sloppy models may be possible, even if the parameters or structure is uncertain, 
if a few critical directions in parameter space are properly constrained. 
Typical ODE models can also require significant mechanistic information.
Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
are emerging paradigms that constrain model structure by the availability of training and validation data. 
These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available, and limits the 
identification challenge. Although not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
However, modeling dynamics using logical models is not straightforward. 
Thus, a potential third approach could be a combination of ODEs and logical models, where mechanistic information was used when available, 
and missing or incomplete knowledge encoded using causal local rules.

@@

Coagulation models in the literature have typically been formulated as systems of nonlinear ordinary differential equations, 
often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined a highly mechanistic model of extrinsic coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
However, most of these previous studies were based upon detailed mechanistic knowledge. 
Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which mechanistic information is difficult or impossible to obtain. 
We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

\subsection*{Formulation of reduced order coagulation models.}
We developed a reduced order coagulation model to test our hybrid kinetic modeling approach (Fig. \ref{fig-network}).
A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
This step loosely represents the initial fluid phase activation of thrombin by activated FXa.
Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) inhibits amplification. 
All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients using an integration rule.  
These control coefficients then modified the rates of model processes at each time step.
Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the influence of components upon a target process. 
Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
However, other integration rules are certainly possible. 
Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact, consisting of only five differential equations. 
Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.

\subsection*{Identification of model parameters using particle swarm optimization.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
We estimated kinetic and control parameters simultaneously from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway. 
The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
the residual. However, not all parameters were varied simultaneously. 
We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.
The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations. The best particle from each generation was used to generate the particle population
for the next generation. We rotated the subproblems, starting with subproblem 1 in the first generation.

@@

The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa and rFVIIa, as we have only a single lumped initiation source (trigger).
Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin being produced was from initiation (Fig. \ref{fig-rFVII}, 1$\times$~trigger).
As we increased the strength of the trigger event, the thrombin peak time and the maximum value of thrombin increased (Fig. \ref{fig-rFVII}, 50$\times$~trigger). 
However, as the trigger strength increased, the thrombin generated quickly decayed.
For example, for the highest trigger strength simulated (50$\times$trigger), 95\% of the thrombin was gone by 20 min after initiation. 
We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.util..
Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig. \ref{fig-flux-analysis}).
The reduced order network includes four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
For normal coagulation, amplification and thrombin inhibition are the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
After 10 min, the dominate reaction has shifted to thrombin inhibition (both ATIII and APC action). In the current proof-of-principle model, 
APC inhibits upstream of amplification thus APC activity will slow the rate of thrombin formation directly from the initiation trigger. 
On the other hand, ATIII inhibits thrombin directly, as well as upstream coagulation factors.
In hemophilia (missing both fVIII and fIX), the amplification reaction does not occur and the only thrombin produced is from initiation (Fig. \ref{fig-flux-analysis}, center).
Initiation is quickly inhibited by APC, and the thrombin level stabilizes and eventually decays because of ATIII activity. 
Lastly, when 50$\times$rFVIIa is used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
initiation mechanisms dominate for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
However, similar to hemophilia alone, without amplification the thrombin signal is quickly extinguished by the combined action of ATIII and APC generated by thrombin.

@@

\section*{Discussion}
Controlling hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
Alternatively, rFVIIa has been used as a universal hemostatic agent to initiate clotting with and without fVIII or fIX inhibitors \cite{Hedner:2008rm}.
Despite its wide adoption, the mechanism and the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}. 
In this study, we formulated a model of the human coagulation cascade and studied thrombin formation in normal and rFVIIa treated and untreated
hemophilic plasma. The model described 193 proteins or protein complexes connected by 301 interactions. 
The coagulation network architecture was based on literature and a previous mathematical model from our laboratory \cite{VARNER2007}.
Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
\emph{in-vitro} studies \cite{BUTENAS2002}.
While the human coagulation cascade is perhaps an ideal model system to develop network analysis tools, not all the model parameters were identifiable given the current training data.
Instead, an \emph{ensemble} of models (N = 437) was estimated using nine cell-based coagulation training sets \cite{ALLEN2007}.
We used this family of consistent coagulation models in all simulations instead of a single best-fit but uncertain model.
Additionally, using sensitivity analysis on the ensemble of models, we identified 
structurally sensitive components of the coagulation architecture as a function of condition.
We also demonstrated that the ensemble robustly constrained model predictions of independent validation sets, despite having many poorly constrained parameters.
Thus, the model ensemble displayed \emph{sloppy} behavior similar to that observed by Sethna and coworkers for other signal transduction networks \cite{Gutenkunst2007}.  
By studying a family of models, we perhaps partially addressed the uncertainty stemming from the many poorly characterized model parameters.
However, many other factors could influence our results. For example, missing structural interactions or biophysical factors such as blood flow could play a large role.
Thus, while the results presented here may be a valuable first step, more studies are required.

@@

While the performance of the proof of principle reduced order model was impressive, there are several questions that should be explored further.
[FINISH]

>>>>

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, but the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model performance is typically 
sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions from sloppy models may be possible, despite parametric uncertainty, if a few critical components are well understood. 
Perhaps a more important issue is that ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
Data-driven \citep{Cirit:2012kx} or logical model approaches \citep{Morris:2011ys} 
are emerging modeling paradigms that constrain model structure by the availability of experimental data. 
These are interesting techniques because the data naturally informs model structure, thereby limiting the 
identification challenge. Although not always mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
However, modeling dynamics using logical models is not straightforward. 
Thus, a potential third approach could be a combination of ODEs and logical models, where mechanistic information was used when available, 
and missing or incomplete knowledge encoded using causal local rules.

@@

Coagulation models in the literature have typically been formulated as systems of nonlinear ordinary differential equations, 
often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
However, most of these previous studies were based upon detailed mechanistic knowledge. 
Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which mechanistic information is difficult or impossible to obtain. 
We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

\subsection*{Formulation of reduced order coagulation models.}
We developed a reduced order coagulation model to test our hybrid kinetic rule based modeling approach (Fig. \ref{fig-network}).
A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
This step loosely represents the initial fluid phase activation of thrombin by activated FXa.
Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) inhibits amplification. 
All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients using an integration rule.  
These control coefficients then modified the rates of model processes at each time step.
Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the influence of components upon a target process. 
Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
However, other integration rules are certainly possible. 
Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact, consisting of only five differential equations. 
Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.

\subsection*{Identification of model parameters using particle swarm optimization.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
We estimated kinetic and control parameters simultaneously from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway. 
The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
the residual. However, not all parameters were varied simultaneously. 
We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations. The best particle from each generation was used to generate the particle population for the next generation. We rotated the subproblems, starting with subproblem 1 in the first generation.

@@

The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa and rFVIIa, as we have only a single lumped initiation source (trigger).
Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin being produced was from initiation (Fig. \ref{fig-rFVII}, 1$\times$~trigger).
As we increased the strength of the trigger event, the thrombin peak time and the maximum value of thrombin increased (Fig. \ref{fig-rFVII}, 50$\times$~trigger). 
However, as the trigger strength increased, the thrombin generated quickly decayed.
For example, for the highest trigger strength simulated (50$\times$trigger), 95\% of the thrombin was gone by 20 min after initiation. 
We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig. \ref{fig-flux-analysis}).
The reduced order network includes four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
For normal coagulation, amplification and thrombin inhibition are the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
After 10 min, the dominate reaction has shifted to thrombin inhibition (both ATIII and APC action). In the current proof-of-principle model, 
APC inhibits upstream of amplification thus APC activity will slow the rate of thrombin formation directly from the initiation trigger. 
On the other hand, ATIII inhibits thrombin directly, as well as upstream coagulation factors.
In hemophilia (missing both fVIII and fIX), the amplification reaction does not occur and the only thrombin produced is from initiation (Fig. \ref{fig-flux-analysis}, center).
Initiation is quickly inhibited by APC, and the thrombin level stabilizes and eventually decays because of ATIII activity. 
Lastly, when 50$\times$rFVIIa is used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
initiation mechanisms dominate for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
However, similar to hemophilia alone, without amplification the thrombin signal is quickly extinguished by the combined action of ATIII and APC generated by thrombin.

@@

%Controlling hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
%Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
%Alternatively, rFVIIa has been used as a universal hemostatic agent to initiate clotting with and without fVIII or fIX inhibitors \cite{Hedner:2008rm}.
%Despite its wide adoption, the mechanism and the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}. 
%The model described the extrinsic activation of thrombin, in the presence of ATIII, with and without the protein C pathway.
%The coagulation network architecture was based on literature and a previous mathematical model from our laboratory \cite{VARNER2007}.
%Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
%\emph{in-vitro} studies \cite{BUTENAS2002}.

%Additionally, using sensitivity analysis on the ensemble of models, we identified 
%structurally sensitive components of the coagulation architecture as a function of condition.
%We also demonstrated that the ensemble robustly constrained model predictions of independent validation sets, despite having many poorly constrained parameters.
%Thus, the model ensemble displayed \emph{sloppy} behavior similar to that observed by Sethna and coworkers for other signal transduction networks \cite{Gutenkunst2007}.  
%By studying a family of models, we perhaps partially addressed the uncertainty stemming from the many poorly characterized model parameters.
%However, many other factors could influence our results. For example, missing structural interactions or biophysical factors such as blood flow could play a large role.
%Thus, while the results presented here may be a valuable first step, more studies are required.
% \cite{ALLEN2007}.

%We used this ensemble of parameters in all subsequent simulations to estimate the uncertainty in model simulations following from uncertainty in the model parameters.

\section*{Discussion}
In this study, we developed a reduced order model of the human coagulation cascade.
We chose coagulation because it is well studied, has a complex architecture, 
and has an abundance of experimental data available for model identification and validation. 
However, coagulation was just a proof-of concept; 
the proposed framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems. 
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
We estimated model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
However, not all of the model parameters were uniquely identifiable, given the training data.
Instead, we estimated an ensemble of likely parameter sets (N = 20)
from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the reduced order hybrid approach produced a surprisingly predictive model, suggesting this approach could potentially 
be used to model other biochemical networks important to human health.

@@

Although the performance of the proof of principle coagulation model was impressive given its dimension, there are several mathematical and biological questions that can be explored further.
First, the choice of max/min integration rules, or the particular form of the transfer functions could be generalized to include other rule types and functions. 
Theoretically, an integration rule is simply a function whose domain is a set of transfer function inputs, and whose range is $[0,1]$.
Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $[0,1]$.
Alternative integration rules such the mean would have different properties which could influence model identification or performance. 
For example, a mean or product integration rule would be differentiable, which allows derivative based optimization approaches to be used. 
The particular form of the transfer function could also be explored. We choose a hill-like function because of its 
prominence in the systems and synthetic biology community. 
However the only mathematical requirement for a transfer function is to map a non-negative continuous or categorical variable into the 
range $[0,1]$. Thus, many other function types are possible.
